---
title: "Functional Data Analysis - Testing the Mean Function"
author: "Gonem Lau"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---


# Introduction

This document presents an analysis aimed at testing the mean function in functional data. The data is generated from a Brownian motion process, and we apply statistical techniques to test hypotheses about the mean function.

# Setup and Initialization

```{r setup-constants}
R_REPLICATIONS <- 200
N_SAMPLES <- 100
K_DESIGN_POINTS <- 20
J_BASES <- 100
L_TIMES <- 250
MC_SAMPLES <- 500
J_TRUNCATED_BASES <- 50

set.seed(2023)
```

# Utility and Mathematical Functions

Defining utility functions for generating Brownian motion eigenvalues, eigenfunctions, and mean functions.

```{r utility-functions}
# Mean functions
zero_mean_fn <- function(time) { 0 }
sinusoidal_mean_fn <- function(time, amplitude = 1, frequency = 1, phase = 0) {
  amplitude * sin(2 * pi * frequency * time + phase)
}

# Error function
error_fn <- function(time, sigma = 0.5) { rnorm(1, mean = 0, sd = sigma) }

# Eigenvalue and eigenfunction calculations for Brownian motion
eigen_value_bm <- function(index) { 1 / (pi^2 * (index - 0.5)^2) }
eigen_values_bm <- function(num_bases) { sapply(seq_len(num_bases), eigen_value_bm) }
eigen_function_bm <- function(index) { function(time) { sqrt(2) * sin(pi * (index - 0.5) * time) } }
eigen_functions_bm <- function(num_bases) { lapply(seq_len(num_bases), eigen_function_bm) }
```

# Data Generation

Simulating design points and generating functional data with noise.

```{r data-generation}
library(MASS)

# Function to generate design points
generate_single_design_points <- function(num_times, is_poisson = FALSE) {
  if (is_poisson) { sort(runif(rpois(1, lambda = num_times), 0, 1)) } 
  else { sort(runif(num_times, 0, 1)) }
}

# Function to generate scores based on eigenvalues
score_generator <- function(num_samples, eigen_values) {
  num_bases <- length(eigen_values)
  cov <- diag(eigen_values)
  MASS::mvrnorm(n = num_samples, mu = rep(0, num_bases), Sigma = cov)
}

# Function for observation generation
generate_observations <- function(observation_functions, num_times, is_poisson = FALSE) {
  lapply(observation_functions, function(f) {
    design_points <- generate_single_design_points(num_times, is_poisson)
    observations <- sapply(design_points, f)
    list(observations = observations, design_points = design_points)
  })
}
```

# Function Estimation

Estimating the functions from our observations using two approaches: linear interpolation and Nadaraya-Watson smoothing.

```{r function-estimation}
# Function generator
function_generator_bm <- function(mean_fn, scores, eigen_fns) {
  apply(scores, 1, function(score_row) {
    function(time) {
      specific <- sum(mapply(FUN = function(eigen_fn, score) eigen_fn(time) * score, eigen_fns, score_row))
      common <- mean_fn(time)
      return(common + specific)
    }
  })
}

# Empirical mean function
empirical_mean_function <- function(functions) {
  function(time) {
    values_at_time <- sapply(functions, function(f) f(time))
    mean(values_at_time)
  }
}

# Observation function generator
observation_function_generator <- function(functions, error_fn) {
  lapply(functions, function(f) {
    function(time) {
      function_value <- f(time)
      error <- error_fn(time)
      function_value + error
    }
  })
}

# Estimators for interpolated and smoothed functions
estimate_interpolated_functions <- function(observations_with_design_points) {
  lapply(observations_with_design_points, function(data) {
    approxfun(data$design_points, data$observations, method = "linear", rule = 2)
  })
}

estimate_smoothed_functions <- function(observations_with_design_points, bandwidth, kernel) {
  lapply(observations_with_design_points, function(data) {
    function(time) {
      distances <- (time - data$design_points) / bandwidth
      weights <- kernel(distances)
      sum(weights * data$observations) / sum(weights)
    }
  })
}
```

# Statistical Testing

Performing statistical testing to assess the behavior of the test statistic under the null hypothesis.

```{r statistical-testing}
# Monte Carlo norm approximation
mc_norm <- function(mc_samples, empirical_mean_fn, mean_fn) {
  sampled_norms <- sapply(mc_samples, function(time) {
    (empirical_mean_fn(time) - mean_fn(time))^2
  })
  mean(sampled_norms)
}

# Test norm calculation
test_norm <- function(num_functions, mc_samples, empirical_mean_fn, mean_fn) {
  num_functions * mc_norm(mc_samples, empirical_mean_fn, mean_fn)
}

# Null hypothesis mean realization
mean_null_realization <- function(eigen_values) {
  random_variables <- rchisq(length(eigen_values), df = 1)
  sum(eigen_values * random_variables)
}

# Monte Carlo quantiles calculation
mc_quantiles <- function(mc_samples, eigen_values, probabilities) {
  simulated_sums <- replicate(mc_samples, mean_null_realization(eigen_values))
  quantile(simulated_sums, probabilities)
}

# Replication of the test norm
replicate_test_norm <- function(...) {
  # Your existing code here
}

# Gaussian Nadaraya-Watson kernel
gaussianNW <- function(obs) estimate_smoothed_functions(obs, 0.1, dnorm)
```


# Sampling and Replication Functions

The following functions are used for sampling time points and replicating the test norm across multiple samples. This is crucial for assessing the behavior of our test statistic under various conditions.

```{r sampling-replication-functions}
# Sampling random times uniformly within a specified range
sample_times <- function(n_samples, min_time, max_time) {
  random_samples <- runif(n_samples, min = min_time, max = max_time)
  sort(random_samples)
}

# Generating a grid of times
grid_times <- function(n_samples, min_time, max_time) {
  seq(min_time, max_time, length.out = n_samples)
}

# Replicating the test norm for each sample
replicate_test_norm <- function(num_points, num_times, time_sampler, num_samples, eigen_values, eigen_functions, function_estimator, mean_fn, error_fn, is_poisson = TRUE, test_mean_fn = NULL) {
  if (is.null(test_mean_fn)) {
    test_mean_fn <- mean_fn
  }

  scores <- score_generator(num_samples, eigen_values)
  functions <- function_generator_bm(mean_fn, scores, eigen_functions)
  observation_functions <- observation_function_generator(functions, error_fn)

  observations <- generate_observations(observation_functions, num_points, is_poisson)
  estimated_functions <- function_estimator(observations)

  empirical_mean_fn <- empirical_mean_function(estimated_functions)
  sampled_times <- time_sampler(num_times, 0, 1)

  test_norm(num_samples, sampled_times, empirical_mean_fn, test_mean_fn)
}
```

# Analysis Execution and Results

Executing the analysis and generating QQ plots to visualize the test statistic behavior under different mean functions.

```{r analysis-execution}
# Parallel processing setup
library(parallel)
num_cores <- parallel::detectCores() * 2 / 3

# Brownian Motion simulation setup
eigen_values <- eigen_values_bm(J_BASES)
eigen_functions <- eigen_functions_bm(J_BASES)

# Calculating theoretical quantiles
probs_seq <- seq(0, 1, by = 0.01)
truncated_eigen_values <- eigen_values[seq_len(J_TRUNCATED_BASES)]
quantiles <- mc_quantiles(MC_SAMPLES, truncated_eigen_values, probs_seq)


zero_linear_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, estimate_interpolated_functions, zero_mean_fn, error_fn), mc.cores = num_cores)
zero_smoothed_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, gaussianNW, zero_mean_fn, error_fn), mc.cores = num_cores)
sinusoidal_linear_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, estimate_interpolated_functions, sinusoidal_mean_fn, error_fn), mc.cores = num_cores)
sinusoidal_smoothed_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, gaussianNW, sinusoidal_mean_fn, error_fn), mc.cores = num_cores)


zero_linear_results <- sapply(quantiles, function(quantile) mean(zero_linear_test_norm_results <= quantile))
zero_smoothed_results <- sapply(quantiles, function(quantile) mean(zero_smoothed_test_norm_results <= quantile))
sinusoidal_linear_results <- sapply(quantiles, function(quantile) mean(sinusoidal_linear_test_norm_results <= quantile))
sinusoidal_smoothed_results <- sapply(quantiles, function(quantile) mean(sinusoidal_smoothed_test_norm_results <= quantile))


zero_linear_results
zero_smoothed_results
sinusoidal_linear_results
sinusoidal_smoothed_results


# Calculate theoretical & empirical quantiles of test_norm_results
zero_linear_empirical_quantiles <- quantile(unlist(zero_linear_test_norm_results), probs = probs_seq)
zero_smoothed_empirical_quantiles <- quantile(unlist(zero_smoothed_test_norm_results), probs = probs_seq)
sinusoidal_linear_empirical_quantiles <- quantile(unlist(sinusoidal_linear_test_norm_results), probs = probs_seq)
sinusoidal_smoothed_empirical_quantiles <- quantile(unlist(sinusoidal_smoothed_test_norm_results), probs = probs_seq)

```


```{r qq-plots}
# QQ Plot for Zero Mean Linearly Interpolated
plot(quantiles, zero_linear_empirical_quantiles, main = "QQ Plot for Zero Mean Linearly Interpolated", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Zero Mean Gaussian NW Smoothed
plot(quantiles, zero_smoothed_empirical_quantiles, main = "QQ Plot for Zero Mean Gaussian NW Smoothed", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Sinusoidal Mean Linearly Interpolated
plot(quantiles, sinusoidal_linear_empirical_quantiles, main = "QQ Plot for Sinusoidal Mean Linearly Interpolated", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Sinusoidal Mean Gaussian NW Smoothed
plot(quantiles, sinusoidal_smoothed_empirical_quantiles, main = "QQ Plot for Sinusoidal Mean Gaussian NW Smoothed", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")
```


```{r analysis-execution-2}
K_DESIGN_POINTS <- 100
small_error_fn <- function(time) error_fn(time, sigma = .1)

zero_linear_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, estimate_interpolated_functions, zero_mean_fn, small_error_fn), mc.cores = num_cores)
zero_smoothed_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, gaussianNW, zero_mean_fn, small_error_fn), mc.cores = num_cores)
sinusoidal_linear_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, estimate_interpolated_functions, sinusoidal_mean_fn, small_error_fn), mc.cores = num_cores)
sinusoidal_smoothed_test_norm_results <- mclapply(seq_len(R_REPLICATIONS), function(i) replicate_test_norm(K_DESIGN_POINTS, L_TIMES, grid_times, N_SAMPLES, eigen_values, eigen_functions, gaussianNW, sinusoidal_mean_fn, small_error_fn), mc.cores = num_cores)


zero_linear_results <- sapply(quantiles, function(quantile) mean(zero_linear_test_norm_results <= quantile))
zero_smoothed_results <- sapply(quantiles, function(quantile) mean(zero_smoothed_test_norm_results <= quantile))
sinusoidal_linear_results <- sapply(quantiles, function(quantile) mean(sinusoidal_linear_test_norm_results <= quantile))
sinusoidal_smoothed_results <- sapply(quantiles, function(quantile) mean(sinusoidal_smoothed_test_norm_results <= quantile))


zero_linear_results
zero_smoothed_results
sinusoidal_linear_results
sinusoidal_smoothed_results


# Calculate theoretical & empirical quantiles of test_norm_results
zero_linear_empirical_quantiles <- quantile(unlist(zero_linear_test_norm_results), probs = probs_seq)
zero_smoothed_empirical_quantiles <- quantile(unlist(zero_smoothed_test_norm_results), probs = probs_seq)
sinusoidal_linear_empirical_quantiles <- quantile(unlist(sinusoidal_linear_test_norm_results), probs = probs_seq)
sinusoidal_smoothed_empirical_quantiles <- quantile(unlist(sinusoidal_smoothed_test_norm_results), probs = probs_seq)

```

```{r qq-plots-2}
# QQ Plot for Zero Mean Linearly Interpolated
plot(quantiles, zero_linear_empirical_quantiles, main = "QQ Plot for Zero Mean Linearly Interpolated Large K Small Sigma", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Zero Mean Gaussian NW Smoothed
plot(quantiles, zero_smoothed_empirical_quantiles, main = "QQ Plot for Zero Mean Gaussian NW Smoothed Large K Small Sigma", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Sinusoidal Mean Linearly Interpolated
plot(quantiles, sinusoidal_linear_empirical_quantiles, main = "QQ Plot for Sinusoidal Mean Linearly Interpolated Large K Small Sigma", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")

# QQ Plot for Sinusoidal Mean Gaussian NW Smoothed
plot(quantiles, sinusoidal_smoothed_empirical_quantiles, main = "QQ Plot for Sinusoidal Mean Gaussian NW Smoothed Large K Small Sigma", xlab = "Theoretical Quantiles", ylab = "Empirical Quantiles", pch = 19)
abline(0, 1, col = "red")
```


# Conclusion

In this analysis, we conducted a comprehensive functional data analysis to test the mean function under different scenarios. Our methodology included generating functional data, estimating functions using linear interpolation and Nadaraya-Watson smoothing, and statistically testing the mean function. 


